---
title: "Full tutorial: from genotypes to a map of genetic range shifts"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 9, fig.height = 5.4,
                      dpi = 110, out.width = "100%", cache = TRUE)
```

This tutorial walks through the full corridoR workflow, step by step, on a
simulated mountain range. It follows the six steps of the analysis in
[Maier et al. (2022, *Heredity*)](https://doi.org/10.1038/s41437-022-00561-x),
which forecast upslope range shifts in the Yosemite toad. The same steps are
shown on the real toad data in the
[Yosemite toad tutorial](https://www.paulmaierresearch.com/software/yosemite-toad-corridors/).
The full run takes about 15 minutes on a laptop; most of that is step 3.

```{r packages, message = FALSE}
library(corridoR)
library(terra)
library(sf)
```

## The simulated range

The example range is modeled loosely on the central Sierra Nevada. A long,
gradual western slope rises from foothills at about 1000 m to a crest above
3400 m near the eastern edge. East of the crest the ground falls away steeply
into a rain shadow. Two river canyons cut the western slope, with walls steep
enough to block movement except near their heads.

Eighty meadows are sampled, eight snowmelt-dependent toads in each. They belong to
four lineages: West and South at low elevation on either side of the southern
canyon, and North and East at high elevation. Lineages at the same elevation
share their environment but not their history.

```{r study-area, echo = FALSE, fig.width = 11, fig.height = 6.25}
ex <- corridor_example()
lc <- c(North = "#ff3939", East = "#0070ff", West = "#000000", South = "#4ce600")
hill <- shade(terrain(ex$dem * 3, "slope", unit = "radians"), terrain(ex$dem, "aspect", unit = "radians"), 35, 315)
dem_cols <- colorRampPalette(c("#5b7f4a", "#9fae6a", "#d8cf98", "#c9b28a", "#ece6dc", "#ffffff"))(80)

layout(matrix(1:2, 1), widths = c(4, 1.25))
par(mar = c(1, 1, 2, 1))
plot(hill, col = grey(0:100 / 100), legend = FALSE, axes = FALSE, main = "Simulated range: meadows and lineages")
plot(ex$dem, col = dem_cols, alpha = 0.8, add = TRUE,
     plg = list(title = "Elevation (m)", x = "bottomleft", horiz = TRUE, cex = 0.7))
plot(st_geometry(ex$sites), add = TRUE, pch = 21, bg = lc[ex$sites$lineage], col = "white", cex = 1.6, lwd = 1.1)
text(c(4000, 4000), c(30500, 13800), c("north canyon", "south canyon"), col = "grey15", cex = 0.7, adj = 0, font = 3)
text(50500, 39000, "crest", col = "grey15", cex = 0.75, font = 3)

# lineage tree from lineage_tmrca: (North, ((East, West), South))
tm <- ex$lineage_tmrca
age <- function(a, b) tm$tmrca[(tm$lineage1 == a & tm$lineage2 == b) | (tm$lineage1 == b & tm$lineage2 == a)]
root <- age("North", "West"); ew <- age("East", "West"); s_node <- age("West", "South")
yy <- c(North = 4, East = 3, West = 2, South = 1)
par(mar = c(4, 0.5, 2, 0.5))
plot(NA, xlim = c(root * 1.12, -root * 0.55), ylim = c(0.5, 4.6), axes = FALSE, xlab = "", ylab = "",
     main = "Lineages", cex.main = 1)
seg <- function(x0, y0, x1, y1, ...) segments(x0, y0, x1, y1, lwd = 3, lend = 1, ...)
y_ew <- mean(yy[c("East", "West")]); y_s <- mean(c(y_ew, yy["South"])); y_root <- mean(c(yy["North"], y_s))
seg(ew, yy["East"], ew, yy["West"]); seg(s_node, y_ew, s_node, yy["South"]); seg(root, y_s, root, yy["North"])
seg(root, y_s, s_node, y_s); seg(s_node, y_ew, ew, y_ew)
for (l in names(yy)) {
  from <- switch(l, North = root, South = s_node, ew)
  seg(from, yy[l], 0, yy[l], col = lc[l])
  points(0, yy[l], pch = 21, bg = lc[l], col = "white", cex = 2)
  text(-root * 0.06, yy[l], l, adj = 0, cex = 0.9)
}
axis(1, at = seq(0, root, by = 200), labels = seq(0, root, by = 200), cex.axis = 0.7)
mtext("Thousand years ago", 1, line = 2.3, cex = 0.7)
```

The simulation behind the genetic data works like this. Toads move most easily
over gentle, forested ground. Migrants leave a meadow for wetter neighbors, and
the drier their own meadow, the stronger that pull. Under the warming scenario
(+4.2 C), snowpack shrinks by a share that falls with elevation, from about 90%
in the foothills to under 20% at the crest, as in climate projections for
Yosemite. Low meadows dry out the most, so the push uphill grows most at low
elevation. The task is to recover that pattern from genotypes and landscape
data alone.

```{r example}
ex <- corridor_example(acc = TRUE)
names(ex)
ex$sites
```

`ex$env` and `ex$env_future` hold the environmental layers now and under
warming. The change in meadow moisture is largest in the foothills and fades
toward the crest:

```{r climate, fig.height = 3.4}
plot(c(ex$env$moisture, ex$env_future$moisture - ex$env$moisture),
     main = c("Meadow moisture index now", "Change under warming"),
     col = hcl.colors(100, "viridis"), axes = FALSE, mar = c(1, 1, 2, 5))
```

## Step 1. Genetic differentiation and the direction of migration

`read_genotypes()` reads STRUCTURE, GENEPOP, VCF, GenAlEx, FSTAT and PLINK
files and returns allele counts per population. `as_popcounts()` does the same
for `adegenet` and `vcfR` objects. The example ships the same SNPs in three
formats:

```{r read}
pc <- read_genotypes(ex$files$structure)
pc
pc_vcf <- read_genotypes(ex$files$vcf, pop = ex$files$popmap)   # VCF needs a population map
all.equal(pairwise_fst(pc), pairwise_fst(pc_vcf))
```

Pairwise F<sub>ST</sub> measures differentiation. Directional G<sub>ST</sub> (Sundqvist et al.
2016, as in `diveRsity::divMigrate`) estimates relative migration in each
direction from a hypothetical gene pool shared by each pair; the difference,
emigration minus immigration, is δM. Positive δM means the source sends more
migrants than it receives.

```{r genetics}
fst <- pairwise_fst(pc)
dg <- directional_gst(pc)
gen <- pairwise_table(fst, dg$dM)
head(gen)
```

Ordering the matrices by lineage shows the structure:

```{r matrices, fig.height = 4.4}
ord <- order(match(ex$sites$lineage, c("North", "East", "West", "South")), ex$sites$elevation)
ids <- ex$sites$site[ord]
op <- par(mfrow = c(1, 2), mar = c(1, 1, 2, 4))
image(t(fst[rev(ids), ids]), axes = FALSE, col = hcl.colors(50, "Inferno"), main = "FST")
lim <- max(abs(dg$dM), na.rm = TRUE)
image(t(dg$dM[rev(ids), ids]), axes = FALSE, zlim = c(-lim, lim),
      col = hcl.colors(50, "Blue-Red 3"), main = "dM (row to column)")
par(op)
```

## Step 2. Choosing the most likely migration paths

Corridors start from a resistance surface, which you supply. In the paper,
`ResistanceGA` optimized the resistance of slope and of vegetation moisture,
and eleven blends of the two were compared. The example includes both
components, so we can do the same.

```{r blends}
xy <- st_coordinates(ex$sites); rownames(xy) <- ex$sites$site
pairs <- as.data.frame(t(combn(ex$sites$site, 2))); names(pairs) <- c("from", "to")
pairs <- pairs[sqrt(rowSums((xy[pairs$from, ] - xy[pairs$to, ])^2)) < 15000, ]
nrow(pairs)

blend <- function(w) {
  r <- w * ex$resistance_slope + (1 - w) * ex$resistance_cover
  r[ex$resistance_slope >= 1e6] <- 1e6
  r
}
weights <- seq(0, 1, by = 0.2)
surfaces <- setNames(lapply(weights, blend), paste0("slope_", weights))
```

`rank_resistance()` finds the cost of the cheapest route between each pair of
meadows through every surface (ridgelines and canyon walls at resistance 1e6
cannot be crossed) and ranks the surfaces by a mixed model of F<sub>ST</sub> on that
distance, with source and destination meadow as random effects. Straight-line
distance is added as a baseline.

```{r rank}
rank <- rank_resistance(surfaces, ex$sites, fst, pairs = pairs)
rank
```

Slope-heavy hypotheses fit best: pure slope ranks first, just ahead of the
80:20 blend the simulation used, and every path model beats straight-line
distance. We continue with `ex$resistance`, that 80:20 blend, and draw the
least cost paths. Path length is measured over the terrain surface.

```{r paths}
tr <- make_transition(ex$resistance, barrier = 1e6)
paths <- least_cost_paths(tr, ex$sites, pairs = pairs, dem = ex$dem)
paths$fst <- fst[cbind(paths$from, paths$to)]
lc <- c(North = "#ff3939", East = "#0070ff", West = "#000000", South = "#4ce600")
hill <- shade(terrain(ex$dem * 3, "slope", unit = "radians"), terrain(ex$dem, "aspect", unit = "radians"), 35, 315)
plot(hill, col = grey(0:100 / 100), legend = FALSE, axes = FALSE, mar = c(0.5, 0.5, 2, 1),
     main = "Least cost paths, colored by FST")
pc_col <- hcl.colors(50, "Inferno")[cut(paths$fst, 50)]
plot(st_geometry(paths), col = pc_col, lwd = 1.2, add = TRUE)
plot(st_geometry(ex$sites), pch = 21, bg = lc[ex$sites$lineage], col = "white", cex = 1.3, add = TRUE)
```

## Step 3. Corridors and environmental extraction

A toad does not walk a one-pixel line. A least cost corridor keeps the cheapest
share `q` of the landscape between two meadows and weights each kept cell from
1 on the best route to 0 at the corridor edge:

```{r corridor, fig.height = 3.2}
w <- lcc_weights(ex$acc, "M20", "M60", q = c(0.005, 0.02, 0.05))
plot(trim(w), col = hcl.colors(50, "Inferno"), axes = FALSE, nc = 3,
     main = c("q = 0.005", "q = 0.02", "q = 0.05"))
```

`corridor_extract()` summarizes every environmental layer inside every
corridor, weighting cells by corridor weight; with path buffers it uses the
share of each cell inside the buffer. Present and future layers are extracted
together. We try three bandwidths: a 400 m buffer and two corridor widths.

```{r extract}
both <- c(ex$env, ex$env_future)
names(both) <- c(names(ex$env), paste0(names(ex$env), "_f"))
bands <- list(
  lcp_400m = corridor_extract(both, pairs, buffers = path_buffers(paths, 400)[[1]], progress = FALSE),
  lcc_0.01 = corridor_extract(both, pairs, acc = ex$acc, q = 0.01, progress = FALSE),
  lcc_0.05 = corridor_extract(both, pairs, acc = ex$acc, q = 0.05, progress = FALSE)
)
head(bands$lcc_0.05[, 1:6])
```

## Step 4. A bandwidth for each group of features

Climate may act over a broad swath of landscape, while slope matters right
along the route. `select_bandwidth()` fits a random forest of F<sub>ST</sub> on each
feature group under each bandwidth and keeps the one with the lowest
cross-validated error.

```{r bandwidth}
groups <- list(climate = c("snowpack", "runoff", "summer_temp"), wetness = "moisture",
               terrain = "slope", cover = "forest")
sel <- select_bandwidth(bands, fst[cbind(pairs$from, pairs$to)], groups, num.trees = 300)
sel
best <- attr(sel, "best")
best
```

## Step 5. Models of F<sub>ST</sub> and δM

The model data hold one row per *directed* pair. Corridor features are the
same in both directions, so `both_directions()` repeats them. `site_contrast()`
adds at-site features: the difference between the source and destination
meadow. Path length and the lineage effect (`lineage_cross()`: time since two
lineages split, zero within a lineage) go in as they are.

```{r modeldata}
pick <- function(suffix = "") {
  out <- pairs
  for (g in names(groups)) out[groups[[g]]] <- bands[[best[[g]]]][paste0(groups[[g]], suffix)]
  out$path_length <- paths$length
  both_directions(out)
}
now <- pick(); future <- pick("_f")
now <- cbind(now, site_contrast(ex$sites, now, env = ex$env)[, -(1:2)])
future <- cbind(future, site_contrast(ex$sites, now, env = ex$env_future)[, -(1:2)])
lin <- st_drop_geometry(ex$sites)[, c("site", "lineage")]
now$LineageCross <- future$LineageCross <- lineage_cross(now$from, now$to, lin, ex$lineage_tmrca)
y <- gen[match(paste(now$from, now$to), paste(gen$from, gen$to)), ]
model_groups <- c(groups, list(climate.at = paste0(groups$climate, ".at")))
```

`fit_connectivity()` runs a PCA within each feature group, tunes a Cubist model
(rule-based trees with a linear model in every leaf, which extrapolate better
than random forests), drops collinear features by importance, and refits.

```{r models}
m_fst <- fit_connectivity(now, y$fst, model_groups, committees = c(1, 10, 50), neighbors = c(0, 5))
m_dm <- fit_connectivity(now, y$dM, model_groups, committees = c(1, 10, 50), neighbors = c(0, 5))
m_fst
m_dm
```

```{r fit, fig.height = 3.6}
op <- par(mfrow = c(1, 2), mar = c(4, 4, 2, 1))
plot(m_fst$observed, m_fst$cv_pred, pch = 16, col = "#2a7f9e44", xlab = "Observed FST",
     ylab = "Cross-validated prediction", main = "FST"); abline(0, 1)
plot(m_dm$observed, m_dm$cv_pred, pch = 16, col = "#2a7f9e44", xlab = "Observed dM",
     ylab = "Cross-validated prediction", main = "dM"); abline(0, 1)
par(op)
```

F<sub>ST</sub> is explained mostly by corridor features and path length, δM mostly by
contrasts between meadows, as in the Yosemite toad:

```{r importance, fig.height = 4}
op <- par(mfrow = c(1, 2), mar = c(4, 9, 2, 1))
barplot(rev(head(m_fst$importance, 10)), horiz = TRUE, las = 1, cex.names = 0.75,
        col = "#2a7f9e", main = "FST", xlab = "Relative importance")
barplot(rev(head(m_dm$importance, 10)), horiz = TRUE, las = 1, cex.names = 0.75,
        col = "#c8553d", main = "dM", xlab = "Relative importance")
par(op)
```

## Step 6. Forecasting under climate change

`predict()` projects future climate onto the present-day PCA loadings and
predicts each pair now and under warming.

```{r predict}
change <- data.frame(from = now$from, to = now$to,
                     fst = predict(m_fst, future) - predict(m_fst, now),
                     dM = predict(m_dm, future) - predict(m_dm, now))
summary(change[, c("fst", "dM")])
```

`map_pairwise()` spreads each pair's change over its corridor and averages
overlapping corridors. For δM it also sums direction vectors, each pointing
from source to destination and weighted by the size of the change, to give a
net direction in every cell. `plot_shift()` draws the result: blue where change
is small, red where it is large.

```{r map-dm, fig.height = 6}
shift <- map_pairwise(data.frame(change[, 1:2], value = change$dM), ex$acc, ex$sites, progress = FALSE)
plot_shift(shift, dem = ex$dem, sites = ex$sites, col_sites = lc[ex$sites$lineage],
           main = "Projected shift in net migration", legend_title = "|change in dM|")
```

The pressure falls on the lowest meadows. On both sides of each canyon, long
arrows run east, upslope, from the foothills; they shorten through the middle
elevations and all but vanish near the crest, the refuge. East of the crest
the arrows turn and climb back west up the escarpment. The canyons stand out as
gaps that corridors cannot cross. This is the pattern built into the
simulation, recovered from genotypes and landscape data.

The F<sub>ST</sub> model maps present-day connectivity the same way. Scaling predicted
F<sub>ST</sub> so that 1 is the best-connected pair, and averaging over corridors, shows
corridors of high flow and the pinch points between them:

```{r map-conn, fig.height = 6}
fst_now <- predict(m_fst, now)
conn <- data.frame(from = now$from, to = now$to, value = (max(fst_now) - fst_now) / diff(range(fst_now)))
conn_map <- map_pairwise(conn, ex$acc, ex$sites, direction = FALSE, absolute = FALSE, progress = FALSE)
plot(hill, col = grey(0:100 / 100), legend = FALSE, axes = FALSE, mar = c(0.5, 0.5, 2, 6),
     main = "Present-day connectivity")
plot(conn_map$value, col = hcl.colors(50, "Plasma"), alpha = 0.7, add = TRUE,
     plg = list(title = "Connectivity"))
plot(st_geometry(ex$sites), pch = 21, bg = lc[ex$sites$lineage], col = "white", cex = 1.3, add = TRUE)
```

## Using your own data

* **Genotypes:** any format `read_genotypes()` reads, or matrices of F<sub>ST</sub> and
  δM from other software.
* **Sites:** an `sf` layer of points or polygons whose ID column matches the
  population names.
* **Resistance:** a raster per hypothesis, built by hand from layers such as
  slope or vegetation, or optimized with `ResistanceGA`; `rank_resistance()`
  compares them as in step 2.
* **Environment:** any stack of rasters, with future versions under the same
  layer names.
* **Lineages:** a table of site lineages and a table of divergence times for
  `lineage_cross()`.
* **Scale:** `accumulated_cost()` can write to disk, and corridors are built one
  pair at a time, so large landscapes need time rather than memory.

## Citation

If you use corridoR, please cite both papers:

Maier PA, Vandergast AG, Ostoja SM, Aguilar A, Bohonak AJ (2022) Landscape
genetics of a sub-alpine toad: climate change predicted to induce upward range
shifts via asymmetrical migration corridors. *Heredity* 129:257-272.
<https://doi.org/10.1038/s41437-022-00561-x>

Maier PA (2027) corridoR: An R package for forecasting genetic range shifts
along migration corridors. *Methods in Ecology and Evolution*, in review.

## Other references

Sundqvist L, Keenan K, Zackrisson M, Prodohl P, Kleinhans D (2016)
Directional genetic differentiation and relative migration. *Ecology and
Evolution* 6:3461-3475.
